### 2022-11-30 chapter1/chapter1.tex
Updated realignment.py figures to reflect new alignments

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions. More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment.

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/merged.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 23D9 \textbf{(A)}, 01E0 \textbf{(B)}, and 07C3 \textbf{(C)} before and after refinement.}
\label{fig:realignment}
\end{figure}

>>>>

\subsection*{Alignment refinement}
Though the pipeline’s quality control measures ensure a high degree of overall sequence identity between members of an orthologous group, some sequences contain long ``poorly supported'' segments which have no homology to most or any other sequences in the alignment. Since most common multiple sequence alignment algorithms assume the sequences are largely homologous, these segments are sometimes ``over-aligned'' by forcing them into alignment where chance sequence similarities occur. Typically, these segments remain contiguous, so the alignments alternate between short runs of columns with few or no gaps and large gap-rich regions (Fig.~\ref{fig:realignment}A-B, left). More rarely, when long poorly supported segments are adjacent to a long gap in the same sequence, the two are interlaced, yielding long gaps interrupted by short segments of spurious alignment (Fig.~\ref{fig:realignment}C, left).

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, with higher values yielding a stronger correction. While values above 0.8 completely eliminate over-alignment and successfully align highly conserved regions, the alignment as a whole is severely degraded, as even homologous sequences with a small amount of divergence are separated by gaps. Thus, in our pipeline orthologous groups were aligned in two stages. In the first, the sequences were aligned with a strong correction of 0.7. Highly conserved regions were identified, which divided the alignment into a complementary set of diverged regions. The sequences in each of these regions were then extracted and aligned separately with a more conservative value for \textit{a\textsubscript{max}} of 0.4. The resulting ``sub-alignments'' were ``stitched'' back into their positions in the original alignment. By defining highly conserved ``anchor'' regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}, right).

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/merged.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
\label{fig:realignment}
\end{figure}
